# Crime trends

## Homicide

Homicide includes five legal crime labels: murder, manslaughter, manslaughter under mitigating circumstances, infanticide, and causing of death in the context of an assault offence. In Finnish penal law, the last mentioned category refers to incidents in which the violence was intentional, while the result of death was not. Together, these crime rubrics correspond to the definition of homicide used in the European Homicide Monitor standard [@granath2012], capturing lethal and intentional interpersonal violence.

In the last ten years, the number of these offences has varied between 116 and 74. In 2022, the homicide rate in Finland was 1.49 per 100 000 population.

```{r, include=FALSE}
library(gapminder)
library(flexdashboard)
library(foreign)
library(haven)
library(dplyr)
library(rio)
library(tidyverse)
library(dplyr)
library(writexl)
library(lubridate)
library(compare)
library(arsenal)
library(diffdf)
library(plotly)
library(stats)
library(tibble)
#library(VarDef)
library(gtools)
library(data.table)
library(plyr)
library(readxl)
library(plotly)
library(janitor)
library(hrbrthemes)
library(bslib)
library(kableExtra)
library(gt)


table1 <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx", sheet="table1")

#knitr::kable(table1)
```

**Table 1** Homicide offences reported to the police, Finland 2013--2022 (SOCM.)

|                          | 2013 | 2014 | 2015 | 2016 | 2017 | 2018 | 2019 | 2020 | 2021 | 2022 |
|----------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| Homicides                | 107  | 112  | 116  | 96   | 74   | 95   | 83   | 95   | 94   | 83   |
| Homicide per 100,000 pop | 1.96 | 2.05 | 2.11 | 1.74 | 1.34 | 1.72 | 1.50 | 1.72 | 1.69 | 1.49 |

Figure 1 shows the homicide trend based on cause-of-death statistics from 1955, using 5-year periods, with the year 2020 also given. The trend shows a high homicide rate period from 1970 to the beginning of the 1990s, after which a consistent homicide drop took place. The most recent observation year suggests an end to the drop among male victims. Note that infanticide, defined as including victims less than one year old, has not been included in Figure 1.

```{r, echo=FALSE}

figure1_h <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx",sheet="figure1")

fig2_h <- figure1_h %>% 
  pivot_longer(
    cols = "1955-59":"2020", 
    names_to = "years", 
    values_to =  "rate", 
  )

#cols = starts_with("19"), 

fig22_h <- fig2_h %>% ggplot(aes(x=years, y=rate, colour=group, group=group)) +
  geom_line() + 
  geom_point()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5)) +
  theme(axis.title.y=element_blank()) +
  theme(axis.title.x=element_blank()) +
   theme(axis.text.x=element_text(size=8)) 

```

```{r, echo=FALSE}
ggplotly(fig22_h) %>% layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))   

```

**Figure 1** Homicide fatality rate pare 100 000 population, Finland 1955-2020. Victims younger than 1 year old have not been included.

```{r, echo=FALSE}
#table2 <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx",sheet="table2")

#| tbl-colwidths: [20,10]

# table2_2 <- knitr::kable(table2,"html")

#knitr::kable(table2, format = "html", table.attr = "style='width:80%;'")  %>% 
# kable_paper(bootstrap_options = "striped", full_width = F)
  # kableExtra::kable_styling()

#table2 %>% gt() %>% as_raw_html()

```

Finnish homicide is still characterised by a high proportion of incidents involving drunken marginalised males. However, the presence of alcohol drunkenness in homicide has decreased since the 1990s [@lehti2014] . During the last 10 years, the proportion of offenders under the influence of alcohol has remained rather stable whereas the role of drug use has increased. Finnish homicide trends have been extensively analysed from long duration change [@lehti2020] to analyses of the 1970s homicide boom [@kivivuori2003] and the post-1990s homicide drop in Finland [@lehti2014] and in the European context [@suonp채채2022] .

## Assault

Assaults include three crime rubrics: assault, aggravated assault and petty assault. The recorded number of these offences was 38,231 in 2012, and 34,123 in 2021. The number of reported offences has been stable since 2014, with approximately 6 offences per 1000 population. Numbers and rates were slightly higher in the first two years of the series (Table 2).

**Table 2** Assault offences reported to the police, Finland 2012-2021 (SOCM.)

|                          | 2012   | 2013   | 2014   | 2015   | 2016   | 2017   | 2018   | 2019   | 2020   | 2021   |
|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| Assault offences         | 38,231 | 35,515 | 32,928 | 33,661 | 33,769 | 33,535 | 33,639 | 33,849 | 33,285 | 34,123 |
| Assaults per 100,000 pop | 7.05   | 6.51   | 6.02   | 6.13   | 6.14   | 6.08   | 6.10   | 6.13   | 6.01   | 6.15   |

National victim surveys largely corroborate this stable trend. Assault victimisation has been very stable over the 10-year period.\

```{r, echo=FALSE}

figure1_a <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx",sheet="figure2")

fig2_a <- figure1_a %>% 
  pivot_longer(
    cols = "2012":"2021", 
    names_to = "year", 
    values_to =  "rate", 
  )

#cols = starts_with("19"), 

fig22_a <- fig2_a %>% ggplot(aes(x=year, y=rate, colour=violence, group=violence)) +
  geom_line() + 
  geom_point()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5)) +
  theme(axis.title.y=element_blank()) +
  theme(axis.title.x=element_blank()) +
   theme(axis.text.x=element_text(size=8)) 

```

```{r, echo=FALSE}
ggplotly(fig22_a)  %>% layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))   

```

**Figure 2** Violence victims, % of 15--74-year-olds, 2012-2021, Finland. (FCVS).

The year 2020 shows a violence dip (Figure 2). The temporary decrease could probably be explained by COVID-19 related lockdowns and closures, as bars and restaurants were closed for a period in the spring of 2020. The most recent reading (2021) saw restoration of the standard level of risk. This temporary drop is not seen in police statistics. During the period 1980-2009, when intermittent victim surveys were conducted, Finnish victimisation trends and trends of recorded offences were often divergent, a discrepancy explained by an increasing propensity of victims to report crimes to the police. During the last decade, reporting propensity has saturated to a stable level, a partial explanation why survey and register sources yielded similar trends in assault offences.

## Sexual violence

Sexual violence here comprises rape offences (rape and aggravated rape, including attempts) and child sexual abuse (CSA; child sexual abuse and aggravated child sexual abuse, including attempts) The number of rape offences reported to the police has increased in the past 10 years (Table 3). In 2021 the number increased by more than a quarter (28%) from the previous year. In the longer-term, the rise can partly be explained by legislative changes and victim awareness.

**Table 3** Rape offences reported to the police, Finland 2012-2021 (SOCM)

|                               | 2012  | 2013 | 2014  | 2015  | 2016  | 2017  | 2018  | 2019  | 2020  | 2021  |
|----------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| Rape offences                 | 1,009 | 975  | 1,009 | 1,043 | 1,160 | 1,245 | 1,393 | 1,477 | 1,450 | 1,851 |
| Rape offences per 100,000 pop | 1.86  | 1.79 | 1.84  | 1.90  | 2.11  | 2.26  | 2.52  | 2.67  | 2.62  | 3.34  |

Figure 3 contrasts recorded rapes with victim self-reports on sexual violence. Note that the measurement units are different: recorded offences are measured as offences per 10,000 persons, while the sexual victimisation scale is measured as percentages (much more prevalent). There is some consistency in the trends, even though the rise in the results of FCVS is not statistically significant on an annual level.

```{r, echo=FALSE}

figure1_r <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx",sheet="figure3")

fig2_r <- figure1_r %>% 
  pivot_longer(
    cols = "2012":"2021", 
    names_to = "year", 
    values_to =  "rate", 
  )

#cols = starts_with("19"), 

fig22_r <- fig2_r %>% ggplot(aes(x=year, y=rate, colour=`offences/victims`, group=`offences/victims`)) +
  geom_line() + 
  geom_point()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5)) +
  theme(axis.title.y=element_blank()) +
  theme(axis.title.x=element_blank()) +
  theme(axis.text.x=element_text(size=8)) 
```

```{r, echo=FALSE}
ggplotly(fig22_r) %>% layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))    

```

**Figure 3** Reported rape offences per 10 000 persons (SOCM) and percentage of sexual violence victims among 15--74-year-olds (FCVS), Finland 2012-2021.

In the past ten years, the number of child sexual abuse offences reported to the police has varied annually as the number is dependent on crime series and the way they have been recorded (Table 4). In the last few years, the number has been growing and in 2021, 2089 cases of child sexual abuse were reported to the police. Some of the increase can be explained by changes in law and reporting propensities but the increase in recent years cannot fully be explained by these changes.

**Table 4** Rape offences reported to the police, Finland 2012-2021 (SOCM)

|                                          | 2012  | 2013  | 2014  | 2015  | 2016  | 2017  | 2018  | 2019  | 2020  | 2021  |
|--------------|------|------|------|------|------|------|------|------|------|------|
| Child sexual abuse offences              | 1,542 | 1,634 | 1,389 | 1,217 | 1,218 | 1,168 | 1,373 | 1,709 | 1,770 | 2,089 |
| Child sexual abuse offences per 10,000 p | 2.84  | 3     | 2.54  | 2.22  | 2.21  | 2.12  | 2.49  | 3.09  | 3.2   | 3.77  |

The Finnish child victim survey (conducted in the years 1988, 2008 and 2013) and School Health Promotion Study (comparable question in 2019) asked ninth graders (age approx. 15) about their sexual experiences with persons at least 5 years older than them. Figure 4 examines trends in self-reported crime in relation to the crimes reported to the police. Note the different measurement units. The initial trend suggests a decreasing trend with increasing official control attention. The latest trends in both data resources have been increasing. Publication of a new measurement is expected soon.

```{r, echo=FALSE, warning=FALSE}

figure1_s <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx",sheet="figure4")

#fig22_s <- figure1_s %>%
#ggplot(aes(x=year)) + 
 # geom_line(aes(y = `Reported child sexual abuse offences per 10000 p`), color = "darkred") +  geom_point(aes(y = `Sexual experiences of ninth graders with adults, %`)) +
  #theme_classic() +
  #theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5)) +
  #theme(axis.title.y=element_blank()) +
  #theme(axis.title.x=element_blank()) +
   #theme(axis.text.x=element_text(size=8)) 


fig22_s2 <- plot_ly(figure1_s, x = ~year, y = ~`Reported child sexual abuse offences per 10000 p`, name = "Reported child sexual abuse offences per 10000 p", type = 'scatter', mode = 'lines') %>% layout(xaxis = list(title = " "), yaxis = list(title = " ")) %>%  layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))    

fig22_s2 <- fig22_s2 %>% add_trace(y = ~`Sexual experiences of ninth graders with adults, %`, name = "Sexual experiences of ninth graders with adults, %", mode = 'markers') 

fig22_s2

```

**Figure 4** Reported child sexual abuse offences per 10000 persons (SOCM) and percentage of ninth graders with sexual experiences with adults **CVS** [@fagerlund2014] ; [@finnish2019].

## Theft

Theft offences include petty theft, theft, and aggravated theft. The number of thefts reported to the police has remained relatively stable in absolute terms and per population. In 2020, the year of COVID-19 lockdowns, the number of reported theft offences were above average. Within the general theft category, aggravated theft decreased from 3,248 in 2012 to 2,147 in 2021.

|                                       | 2012    | 2013    | 2014    | 2015    | 2016    | 2017    | 2018   | 2019    | 2020    | 2021    |
|----------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| Reported theft offences               | 138,750 | 141,395 | 143,141 | 138,571 | 131,432 | 125,431 | 122741 | 126,906 | 139,121 | 127,791 |
| Reported theft offences per 1,000 pop | 25.6    | 25.9    | 26.2    | 25.3    | 23.9    | 22.8    | 22.2   | 23      | 25.1    | 23      |

The Finnish Crime Victim Survey captures the percentage of persons whose personal property was stolen outside the home (Figure 5). This type of victimisation decreased during the last decade. The same decreasing trend can also be seen in vehicle theft, as captured by the FCVS.

```{r, echo=FALSE}

figure1_th <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx",sheet="figure5")

figure2_th <- figure1_th %>% 
  pivot_longer(
    cols = "2013":"2021", 
    names_to = "year", 
    values_to =  "rate", 
  )


figure22_th <- figure2_th %>% ggplot(aes(x=year, y=rate, colour=theft, group=theft)) +
  geom_line() + 
  geom_point()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5)) +
  theme(axis.title.y=element_blank()) +
  theme(axis.title.x=element_blank()) +
  theme(axis.text.x=element_text(size=8)) 
```

```{r, echo=FALSE}
ggplotly(figure22_th) %>% layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))    

```

**Figure 5** Theft of personal property outside home, per cent of the 15--74-year-old population (FCVS), contrasted with reported theft offences per 100 per-sons (SOCM).

The declining trend of theft against persons and households reflects a drop in the theft of vehicles and other physical property. Identity theft and consumer fraud have increased during the last decade, as observed in both administrative statistics and the FCVS survey system (Figure 6).

```{r, echo=FALSE}

figure1_the <- read_excel("Z:/RT-katsaus/crime-trends-in-finland/data_eng.xlsx",sheet="figure6")

figure2_the <- figure1_the %>% 
  pivot_longer(
    cols = "2013":"2021", 
    names_to = "year", 
    values_to =  "rate", 
  )


figure22_the <- figure2_the %>% ggplot(aes(x=year, y=rate, colour=theft, group=theft)) +
  geom_line() + 
  geom_point()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5)) +
  theme(axis.title.y=element_blank()) +
  theme(axis.title.x=element_blank()) +
  theme(axis.text.x=element_text(size=8)) 
```

```{r, echo=FALSE}
ggplotly(figure22_the) %>% layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))    

```

It should be noted that theft offences reported to the police include crimes against businesses and institutions, a quantitatively major crime target. According to the Finnish Crimes against Businesses Survey, theft victimisation against the retail sector remained stable between 2010 and 2018, with some increase in offences committed by unknown persons [@saarikkom채ki2019] . Based on this, it seems that the decreasing trend in theft has benefitted private persons and households more than the businesses in the retail sector.

## Economic crime

Economic crime covers in this instance tax offences, accounting offences and bankruptcy offences. The number of police-recorded economic crimes has decreased considerably from 2019, which marked the highest total numbers of economic crimes in ten years. In 2021, police registered 2,440 tax offences, accounting offences and bankruptcy offences. This means the number of economic offences decreased by 18 per cent from the previous year and 28 per cent from 2019.

|                                      | 2012  | 2013  | 2014  | 2015  | 2016  | 2017  | 2018  | 2019  | 2020  | 2021  |
|------------|------|------|------|------|------|------|------|------|------|------|
| Economic offences                    | 2,243 | 2,281 | 2,666 | 2,685 | 2,954 | 3,129 | 3,159 | 3,383 | 2,983 | 2,440 |
| Economic offences per 10 000 persons | 4.13  | 4.18  | 4.87  | 4.89  | 5.37  | 5.68  | 5.72  | 6.12  | 5.39  | 4.40  |

Before the decrease in reported economic offences, the number of offences had risen continuously since 2010. The decrease from the peak of 2019 has mostly been due to the decrease of tax offences, just as the increase in 2010--2019 was due to the rising numbers of tax offences. Accounting of-fences and bankruptcy offences have remained fairly stable. The observed decrease in tax offences has likely been caused by the changing emphasis in resource allocation within the field of investigation of economic offences and risk-based tax audits.

As regards economic crime, no survey-based data are available to complement the above trend descriptions based on police statistics. In the 2018 Crimes against Businesses Survey (FCBS), the firms were asked to report if their competitors had used illegal means in business activities in relation to employees, such as illegal working conditions and terms. In the hotel and restaurant sector, eight per cent of the firms had perceived such illegal competition, whereas the same percentage in retail sector was 2 per cent [@saarikkom채ki2019]. However, the trend remains unspecified as there has not been repeated survey after 2018.
