# Youth crime

```{r, include=FALSE}
library(gapminder)
library(flexdashboard)
library(foreign)
library(haven)
library(dplyr)
library(rio)
library(tidyverse)
library(dplyr)
library(writexl)
library(lubridate)
library(compare)
library(arsenal)
library(diffdf)
library(plotly)
library(stats)
library(tibble)
#library(VarDef)
library(gtools)
library(data.table)
library(plyr)
library(readxl)
library(plotly)
library(janitor)
library(bslib)
library(kableExtra)
library(gt)

```

Crime trends among young people are monitored in Finland by means of both criminal justice statistics and surveys. In what follows, we use both. As noted above, these basic sources capture different parts of the seriousness continuum. For serious crime *incidents* such as homicide, serious violence, and robbery, administrative statistics are more reliable than surveys. In contrast, surveys are indispensable sources for the prevalence of young *persons’*  participation in less serious mass crime.

The Finnish criminal justice system uses three age categories of youth, with repercussions on how offenders are punished or treated. Persons under 15 years of age are not criminally culpable. In such cases, the social services deal with the matter. However, the crime is entered into the police information system (and therefore into the SOCM). Age categories 15 to 17 and 18 to 20 have special stipulations that reduce penal severity. To match this logic, we have used the same age categories in our descriptions of crime trends.

Over the last 10 years, theft offences by young people have first decreased substantially, and then increased since the year 2022. Property destruction decreased in the first half of the decade among 15- to 20-year-olds, remaining stable after that. This crime type was stable in the youngest age group. Violence (assault offences) increased in the youngest group and decreased in the 18 to 20 category. The middle category of 15 to 17 showed a nonlinear trend: first a decrease and, after 2018, a moderate increase in police-recorded assaults. The increase in the group of less than 15-year-olds was drastic: the number of offences per 10 000 population increased from 11 in 2012 to 25 in 2021.

The Finnish Self-Report Delinquency Study (FSRD) is a nationally representative crime survey among 15–16-year-olds [@suonpaa_nuorten_2024; @kaakinen_school_2022]. Following the logic of this report, it is of some interest to compare the FSRD survey-based findings with the police-recorded crime trends in the age category 15 to 17. These comparisons are shown in Figures 7a, 7b, and 7c below. Note that the police-recorded offending figure is offence-based – meaning that the same individual can appear multiple times in the same dataset – while the survey observations are person-based. Also, the vertical scales differ: the scale is percentages for the survey and rates per 1000 persons for the administrative statistic. Overall, the two sources largely agree on the decrease in offending in the early 2010s. However, the trends in recent years seem to diverge. The increase in theft offences is observed in both datasets whereas assault offenses increased only in the administrative data.  

Based on police-recorded offences, the crime trends have been quite similar in the 18 to 20 and 15 to 17 age groups (decrease and levelling out). In the youngest group, the trends are different and show an increase in recorded violence.

```{r, echo=FALSE, warning=FALSE}

figure1_y7 <- read_excel("C:/LocalData/ikolttol/RTkatsaus_eng/data_eng.xlsx",sheet="figure7")

fig22_y7 <- plot_ly(figure1_y7, x = ~year, y = ~`Reported offences per 1000 pop (SOCM)`, name = "Reported offences per 1000 pop (SOCM)", type = 'scatter', mode = 'lines') %>% layout(xaxis = list(title = " "), yaxis = list(title = " ")) %>%  layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))    

fig22_y7 <- fig22_y7 %>% add_trace(y = ~`Last year prevalence, % (FSRD)`, name = "Last year prevalence, % (FSRD)", mode = 'markers') 

fig22_y7

```

**Figure 7** Theft offences by young persons: comparison of police recorded offences (15--17-year-olds) (SOCM) and self-reported prevalence rates (15--16-year-olds). Finland 2012--2024 (FSRD).\

```{r, echo=FALSE, warning=FALSE}

figure1_y8 <- read_excel("C:/LocalData/ikolttol/RTkatsaus_eng/data_eng.xlsx",sheet="figure8")


fig22_y8 <- plot_ly(figure1_y8, x = ~year, y = ~`Reported offences per 1000 pop (SOCM)`, name = "Reported offences per 1000 pop (SOCM)", type = 'scatter', mode = 'lines') %>% layout(xaxis = list(title = " "), yaxis = list(title = " ")) %>%  layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))    

fig22_y8 <- fig22_y8 %>% add_trace(y = ~`Last year prevalence, % (FSRD)`, name = "Last year prevalence, % (FSRD)", mode = 'markers') 

fig22_y8

```

**Figure 8** Property destruction offences by young persons: comparison of police recorded offences (15--17-year-olds) (SOCM) and self-reported prevalence rates (15--16-year-olds). Finland 2012--2024 (FSRD).

```{r, echo=FALSE, warning=FALSE}

figure1_y9 <- read_excel("C:/LocalData/ikolttol/RTkatsaus_eng/data_eng.xlsx",sheet="figure9")


fig22_y9 <- plot_ly(figure1_y9, x = ~year, y = ~`Reported offences per 1000 pop (SOCM)`, name = "Reported offences per 1000 pop (SOCM)", type = 'scatter', mode = 'lines') %>% layout(xaxis = list(title = " "), yaxis = list(title = " ")) %>%  layout(legend = list(orientation = "h", xanchor = "center", x = 0.5))    

fig22_y9 <- fig22_y9 %>% add_trace(y = ~`Last year prevalence, % (FSRD)`, name = "Last year prevalence, % (FSRD)", mode = 'markers') 

fig22_y9

```

**Figure 9** Assault offences by young persons: comparison of police recorded offences (15--17-year-olds) (SOCM) and self-reported prevalence rates (15--16-year-olds). Finland 2012--2024 (FSRD).

In the category of youths under 15 years of age, police-recorded assaults and robberies have increased over the last seven to eight years [@danielsson2022] . Even though there have been changes in the compulsory reporting of incidents by child protection agencies, the best currently available analyses indicate that the rise of violence in this group cannot be completely explained by increased reporting propensity [@danielsson2022]. The increase is general in urban, suburban, and rural areas. Sharp instruments and taking videos during the incident are increasingly often seen in incidents involving minors [@danielsson2022] .
